---
title: <span style="font-size:22px;">Tidying and Transforming Data</span>
author: "Stephen Jones"
date: "February 24, 2019"
output:
  html_document:
    themes: "paper"
    highlight: "pygments"
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
    df_print: paged
---
***
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
###Create
<style>
div.gold { background-color:#fffcf3; border-radius: 5px; padding: 20px;}
</style>
<div class = "gold">
"Arrival.csv" is created and written to the working directory.
</div>
```{r message=FALSE, warning=FALSE,strip.white=TRUE}
rm(list=ls())

LosAngeles<-c(497,62,694,117)
Phoenix<-c(221,12,4840,415)
SanDiego<-c(212,20,383,65)
SanFrancisco<-c(503,102,320,129)
Seattle<-c(1841,305,201,61)
status<-c("ontime","delayed","ontime","delayed")
airline<-c("ALASKA","ALASKA","AMWEST","AMWEST")
Arrival<-data.frame(airline,status,LosAngeles,Phoenix,SanDiego,SanFrancisco,Seattle)
#create .csv file.
write.csv(Arrival,"Arrival.csv")
```
***
###Read
<style>
div.gold { background-color:#fffcf3; border-radius: 5px; padding: 20px;}
</style>
<div class = "gold">
"Arrival.csv" loads, "kableExtra" renders the table.
</div>
```{r warning=FALSE, message=FALSE, out.width="100%",strip.white=TRUE}
#clear objects
rm(list=ls())

#read .csv, omit numbered column
Arrival<-read.csv("Arrival.csv")[ ,2:8]

#display
require(kableExtra)
arr<-knitr::kable(Arrival,"html",align='lcccccc')%>%
  kable_styling("striped", full_width = F)
arr

```
***
###Transform {.tabset .tabset-fade .tabset-pills}
<style>
div.gold { background-color:#fffcf3; border-radius: 5px; padding: 20px;}
</style>
<div class = "gold">
Table 1 transforms via the gather function from tidyr, setting key as city and value as frequency; each city becomes a row with unique frequency values. Table 2 is created using the spread function, resulting in a column for each airline and pairs of observations for each city. Table 3 and Table 4 display on-time and delayed subsets of data built by the filter command. Various sums are calculated via mutate and sum commands and displayed in Table 5: flight sums by airline (linetotal), city (citytotal), and status (statustot), as well the combinations airline by city (linecitytot), status by city (statcitytot), and airline by status (linestattot).
</div>

####Table 1
```{r warning=FALSE,message=FALSE,out.width="100%",strip.white=TRUE}
require(tidyr)

Arrival3<-gather(Arrival,'LosAngeles','Phoenix','SanDiego','SanFrancisco','Seattle',key='city',value='frequency')
arr3<-knitr::kable(Arrival3,"html",align='lccc')%>%
  kable_styling("striped", full_width = F)%>%
  scroll_box(height="195px")
arr3

```
***
####Table 2
```{r warning=FALSE,message=FALSE,out.width="100%",strip.white=TRUE}

require(dplyr)

Arrival4<-spread(Arrival3,airline,frequency)
arr4<-knitr::kable(Arrival4,"html",align='lccc')%>%
  kable_styling("striped",full_width = F)%>%
  scroll_box(height="195px")
arr4

#Calculate percentage of flights which are delayed.
delayednum<-sum(Arrival4$ALASKA[ which(Arrival4$status=="delayed") ])+sum(Arrival4$AMWEST[ which(Arrival4$status=="delayed") ])
delayedden<-(sum(Arrival4$ALASKA)+sum(Arrival4$AMWEST))
delayedpct<-round((delayednum/delayedden)*100,1)


```
***
####Table 3
```{r warning=FALSE,message=FALSE,out.width="100%",strip.white=TRUE, fig.cap="Subset, flights not delayed"}
Arrival4ontime<-filter(Arrival4,status=="ontime")

arr4ontime<-knitr::kable(Arrival4ontime,"html",align='lccc')%>%
  kable_styling("striped",full_width = F)

arr4ontime
```
***
####Table 4
```{r warning=FALSE,message=FALSE,out.width="100%",strip.white=TRUE, fig.cap="Subset, delayed flights only"}

Arrival4delayed<-filter(Arrival4,status=="delayed")

arr4delayed<-knitr::kable(Arrival4delayed,"html",align='lccc')%>%
  kable_styling("striped",full_width = F)

arr4delayed
```

***
####Table 5
```{r warning=FALSE,message=FALSE,fig.align="center",strip.white=TRUE, fig.cap="Table with calculated totals and means"}

#add frequency total by airline
Arrival3.5<-Arrival3%>%
  group_by(airline)%>%
  mutate(linetotal=sum(frequency))
#add frequency total by city
Arrival3.5<-Arrival3.5%>%
  group_by(city)%>%
  mutate(citytotal=sum(frequency))
#add frequency total by city and airline
Arrival3.5<-Arrival3.5%>%
  group_by(airline,city)%>%
  mutate(linecitytot=sum(frequency))
#add frequency total by status
Arrival3.5<-Arrival3.5%>%
  group_by(status)%>%
  mutate(statustot=sum(frequency))
#add frequency total by status
Arrival3.5<-Arrival3.5%>%
  group_by(status,city)%>%
  mutate(statcitytot=sum(frequency))
#add frequency total by status and city
Arrival3.5<-Arrival3.5%>%
  group_by(status,city)%>%
  mutate(statcitytot=sum(frequency))
#add frequency total by status and airline
Arrival3.5<-Arrival3.5%>%
  group_by(status,airline)%>%
  mutate(linestattot=sum(frequency))

arr3<-knitr::kable(Arrival3.5,"html",align='lccccccccc')%>%
  kable_styling("striped", full_width = T, font_size = 12)%>%
  scroll_box(height="195px")
arr3

```

***
###Analyze {.tabset .tabset-fade .tabset-pills}
<style>
div.gold { background-color:#fffcf3; border-radius: 5px; padding: 20px;}
</style>
<div class = "gold">
Figure 1 shows that 13.3% of ALASKA flights were delayed, while 10.9% of AMWEST flights were delayed, with `r delayedpct`% of flights delayed overall.In Figure 2 we see that flying into San Franciso provedmost troublesome for both airlines, with 28.7% of AMWEST flights into San Francisco delayed and 16.9% of ALASKA flights into San Francisco delayed. Phoenix proved to be the least challanging to schedules with 5.2% and 7.9% of flights delayed for ALASKA and AMWEST, respectively. Figure 3 illustrates flight delay by city; 1 out of every 5 flights into San Francisco are delayed.<br><br>
<strong>Conclusion:</strong> If given the option to choose among these cities for a tightly scheduled flight itinerary, <i>flying into San Francisco should be avoided if possible.</i>
</div>

####Figure 1
```{r warning=FALSE,message=FALSE,strip.white=TRUE,fig.cap="Flight delay percentage by airline",out.width="100%"}

airlinetable<-unique(Arrival3.5[c("airline","status","linetotal","linestattot")])
airlinetable$statpct<-round((airlinetable$linestattot/airlinetable$linetotal)*100,1)
airlinetable2<-airlinetable[ which(airlinetable$status=="delayed"),]

require(ggplot2)
p1<-ggplot(data=airlinetable2, aes(x = airline, y=statpct)) + 
  geom_bar(stat="identity",fill="steelblue")+
  geom_text(aes(label=statpct), vjust=1.6, color="white", size=3.5)+
  ylab("percent delayed")+
  theme_bw()

require(grid)
require(gridExtra)
t1 <- tableGrob(airlinetable[c("airline","status","statpct")], theme=ttheme_default(base_size=10,padding=unit(c(4,4),"mm")),rows=NULL)

grid.arrange(p1, t1, nrow=1)
```


***
####Figure 2
```{r warning=FALSE,message=FALSE,fig.align="center",strip.white=TRUE,fig.cap="Flight delay percentage by airline and city"}
citytable<-unique(Arrival3.5[c("airline","status","city","frequency","linecitytot")])
citytable$statpct<-round((citytable$frequency/citytable$linecitytot)*100,1)
citytable2<-citytable[ which(citytable$status=="delayed"),]

p1.1<-ggplot(data=citytable2, aes(x = airline, y=statpct,group=city)) + 
  geom_bar(stat="identity",fill="steelblue")+
  geom_text(aes(label=statpct), vjust=1.6, color="white", size=3.5)+
  ylab("percent delayed")+
  facet_grid(~city)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45,hjust=1))

t1.1 <- tableGrob(citytable[c("airline","status","city","statpct")], theme=ttheme_default(base_size=8,padding=unit(c(1,3),"mm")),rows=NULL)

grid.arrange(p1.1, t1.1, nrow=1, widths = 2:1)     

```

***
####Figure 3
```{r warning=FALSE,message=FALSE,fig.align="center",strip.white=TRUE,fig.cap="Percent of flights delayed by city"}

totcitytable<-unique(Arrival3.5[c("city","status","citytotal","statcitytot")])
totcitytable$citystatpct<-round((totcitytable$statcitytot/totcitytable$citytotal)*100,1)
totcitytable2<-totcitytable[ which(totcitytable$status=="delayed"),]

p1.2<-ggplot(data=totcitytable2, aes(x = city, y=citystatpct,group=city)) + 
  geom_bar(stat="identity",fill="steelblue")+
  geom_text(aes(label=citystatpct), vjust=1.6, color="white", size=3.5)+
  ylab("percent delayed")+
  theme_bw()

t1.2<-tableGrob(totcitytable[c("city","status","citystatpct")],
                theme=ttheme_default(base_size=10,
                                     padding=unit(c(2,3),"mm")),rows=NULL)

grid.arrange(p1.2, t1.2, nrow=1, widths = 2:1)   

```

